\Large{Adversarial Training against Systematic Uncertainty}

### 2016-10-09 tex/paper.tex
save
toy: break symmetry, hence avoiding Z=0 to be pivotal
add details about network architectures

\begin{figure}
    \includegraphics[width=0.48\textwidth]{figures/training.pdf}
    \vspace{-1cm}
    \caption{Toy example. Training curves for ${\cal L}_f(\theta_f)$, ${\cal L}_r(\theta_f, \theta_r)$
             and ${\cal L}_f(\theta_f) - \lambda {\cal L}_r(\theta_f, \theta_r)$.
             Adversarial training was performed for $200$ iterations, mini-batches of size $M=128$, $K=100$ and $\lambda=50$.}
    \label{fig:toy-training}
\end{figure}

As a guiding toy example, let us consider the binary classification of 2D
data drawn from multivariate gaussians with equal priors, such that
\begin{align}
    x &\sim {\cal N}(\mu=(0,0), \Sigma= \mathbb{1}) \text{ when } Y=0, \\
    x &\sim {\cal N}(\mu=(1,z), \Sigma= \mathbb{1}) \text{ when } Y=1.
\end{align}
The continuous nuisance parameter $Z$ represents in this case our
uncertainty about the exact location of the mean of the second gaussian. Our goal is to
build a classifier $f(\cdot;\theta_f)$ for predicting $Y$ given $X$, but such that
the probability distribution of $f(X;\theta_f)$ is invariant with respect to the
nuisance parameter $Z$.

Assuming a uniform prior over the interval ${\cal Z} = [0;2]$, we start by
generating training data $\{ x_i, y_i, z_i \}_{i=1}^N$, from which we build a
neural network classifier $f$ minimizing ${\cal L}_f(\theta_f)$ without adversarial
training. As shown in the upper plots of Figure~\ref{fig:toy}, the resulting
classifier is not pivotal, as the conditional probability densities of its
decision scores $f(X;\theta_f)$ show large discrepancies between values $z$ of
the nuisance. While not shown here, a classifier trained only from data
generated at the nominal value $Z=0$ would also not be pivotal.

Let us now consider the adversarial training of $f$ against an adversary $r$
implemented as a mixture density network modeling $Z|f(X;\theta_f)$ as a
truncated mixture of four gaussians over ${\cal Z}$. When running
Algorithm~\ref{alg:adversarial-training} as initialized with the classifier $f$
obtained previously, adversarial training effectively reshapes the decision
function so it that becomes almost independent on the nuisance parameter, as
shown in the lower plots of Figure~\ref{fig:toy}. In particular, the conditional
probability densities of the decision scores $f(X;\theta_f)$ are now very
similar to each other, indicating only a small  dependency on the nuisance, as
theoretically expected. The dynamics of adversarial training is illustrated in
Figure~\ref{fig:toy-training}, where the losses ${\cal L}_f$, ${\cal L}_r$ and
${\cal L}_f - \lambda {\cal L}_r$ are evaluated after each iteration of
Algorithm~\ref{alg:adversarial-training}. In the first iterations, we observe
that the global objective ${\cal L}_f - \lambda {\cal L}_r$ is minimized by
making the classifier less accurate, hence the corresponding increase of ${\cal
L}_f$, but which results in a classifier that is more pivotal, hence the
corresponding increase of ${\cal L}_r$ and the total net benefit. As learning
goes, minimizing $E$ then requires making predictions that are more accurate,
hence decreasing ${\cal L}_f$, or that are less dependent on $Z$, hence shaping
$p_{\theta_r}$ so that it becomes more uniform. Indeed, ${\cal L}_f$ eventually
start to slightly decrease, while remaining lower bounded by $\min_{\theta_f}
{\cal L}_f(\theta_f)$ as approximated by the dashed line in the first plot.
Similarly,  ${\cal L}_r$ keeps tending towards the differential entropy $H(p_Z) =
\log(2)$ of the uniform distribution over ${\cal Z}$, as shown by the dashed
line in the second plot. Finally, let us note that the ideal situation of a
classifier that is both optimal and pivotal appears to be unreachable for this
problem, as shown in third plot by the  offset between ${\cal L}_f - \lambda
{\cal L}_r$ and the dashed line approximating $\mathbb{E}_{X} [ H(p_{Y|X}) -
\lambda H(p_Z)]$.

>>>>

\begin{figure}
    \includegraphics[width=0.48\textwidth]{figures/training.pdf}
    \vspace{-1cm}
    \caption{Toy example. Training curves for ${\cal L}_f(\theta_f)$, ${\cal L}_r(\theta_f, \theta_r)$
             and ${\cal L}_f(\theta_f) - \lambda {\cal L}_r(\theta_f, \theta_r)$.
             Adversarial training was performed for $100$ iterations, mini-batches of size $M=128$, $K=100$ and $\lambda=100$.}
    \label{fig:toy-training}
\end{figure}

As a guiding toy example, let us consider the binary classification of 2D
data drawn from multivariate gaussians with equal priors, such that
\begin{align}
    x &\sim {\cal N}((0,0), \begin{bmatrix}
                              1 & -0.5 \\
                              -0.5 & 1
                            \end{bmatrix}) \text{ when } Y=0, \\
    x &\sim {\cal N}((1,Z),  \begin{bmatrix}
                              1 & 0 \\
                              0 & 1
                             \end{bmatrix}) \text{ when } Y=1.
\end{align}
The continuous nuisance parameter $Z$ represents in this case our
uncertainty about the exact location of the mean of the second gaussian. Our goal is to
build a classifier $f(\cdot;\theta_f)$ for predicting $Y$ given $X$, but such that
the probability distribution of $f(X;\theta_f)$ is invariant with respect to the
nuisance parameter $Z$.

Assuming a uniform prior over the interval ${\cal Z} = [0;2]$, we start by
generating training data $\{ x_i, y_i, z_i \}_{i=1}^N$, from which we build a
neural network classifier $f$ minimizing ${\cal L}_f(\theta_f)$ without
adversarial training. The network architecture comprises 2 dense hidden layers of 20
nodes with ReLU activations, followed  by a dense output layer with a single
node with a sigmoid activation. As shown in the upper plots of
Figure~\ref{fig:toy}, the resulting classifier is not pivotal, as the
conditional probability densities of its decision scores $f(X;\theta_f)$ show
large discrepancies between values $z$ of the nuisance. While not shown here, a
classifier trained only from data generated at the nominal value $Z=0$ would
also not be pivotal.

Let us now consider the adversarial training of $f$ against an adversary $r$
implemented as a mixture density network modeling $Z|f(X;\theta_f)$ as a
truncated mixture of four gaussians over ${\cal Z}$. As for $f$, the network
architecture comprises 2 dense hidden layers of 20 nodes with ReLU activations,
but is followed by an output layer of 12 nodes corresponding to the means,
standard deviations and mixture coefficients of the four gaussians. Output nodes
for the mean values come with tanh activations rescaled over ${\cal Z}$, nodes
for the standard deviations with exponential activations to ensure positivity,
while nodes for the mixture coefficients implement the softmax function to
ensure positivity and normalization. When running
Algorithm~\ref{alg:adversarial-training} as initialized with the classifier $f$
obtained previously, adversarial training effectively reshapes the decision
function so it that becomes almost independent on the nuisance parameter, as
shown in the lower plots of Figure~\ref{fig:toy}. In particular, the conditional
probability densities of the decision scores $f(X;\theta_f)$ are now very
similar to each other, indicating only a small  dependency on the nuisance, as
theoretically expected. The dynamics of adversarial training is illustrated in
Figure~\ref{fig:toy-training}, where the losses ${\cal L}_f$, ${\cal L}_r$ and
${\cal L}_f - \lambda {\cal L}_r$ are evaluated after each iteration of
Algorithm~\ref{alg:adversarial-training}. In the first iterations, we observe
that the global objective ${\cal L}_f - \lambda {\cal L}_r$ is minimized by
making the classifier less accurate, hence the corresponding increase of ${\cal
L}_f$, but which results in a classifier that is more pivotal, hence the
corresponding increase of ${\cal L}_r$ and the total net benefit. As learning
goes, minimizing $E$ then requires making predictions that are more accurate,
hence decreasing ${\cal L}_f$, or that are less dependent on $Z$, hence shaping
$p_{\theta_r}$ so that it becomes more uniform. Indeed, ${\cal L}_f$ eventually
start to slightly decrease, while remaining lower bounded by $\min_{\theta_f}
{\cal L}_f(\theta_f)$ as approximated by the dashed line in the first plot.
Similarly,  ${\cal L}_r$ keeps tending towards the differential entropy $H(p_Z) =
\log(2)$ of the uniform distribution over ${\cal Z}$, as shown by the dashed
line in the second plot. Finally, let us note that the ideal situation of a
classifier that is both optimal and pivotal appears to be unreachable for this
problem, as shown in third plot by the  offset between ${\cal L}_f - \lambda
{\cal L}_r$ and the dashed line approximating $\mathbb{E}_{X} [ H(p_{Y|X}) -
\lambda H(p_Z)]$.
